---
title: "Homework 1"
author: "Raul Torres Aragon"
date: "4/2/2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 2M1  

Recall the globe tossing model from the chapter. Compute and plot the grid approximate posterior distribution for each of the following sets of observations. In each case, assume a uniform prior for Pw.  
(1) W,W,W
(2) W,W,W,L
(3) L,W,W,L,W,W,W  

```{r}
# set up grid of potential Pw values
n_grid <- 1e2
pw_grid <- seq( from=0 , to=1 , length.out=n_grid)

# compute likelihood of x water in n tosses
likelihood_1 <- dbinom(x=3, size=3, prob=pw_grid)
likelihood_2 <- dbinom(x=3, size=4, prob=pw_grid)
likelihood_3 <- dbinom(x=5, size=7, prob=pw_grid)

# prior Recall lim(P[Pw, Pw + delta]/delta) as delta -> 0 = 1
prior <- rep(1,n_grid) 

# posterior
get_posterior <- function(prior, likelihood) {
  prior * likelihood / sum(prior * likelihood)
}
posterior_1 <- get_posterior(prior, likelihood_1)
posterior_2 <- get_posterior(prior, likelihood_2)
posterior_3 <- get_posterior(prior, likelihood_3)


par(mfrow = c(1,3))
plot(posterior_1 ~ pw_grid, type="l", xlab="Pw values", ylab="probability", main="W W W")
plot(posterior_2 ~ pw_grid, type="l", xlab="Pw values", ylab="", main="W W W L")
plot(posterior_3 ~ pw_grid, type="l", xlab="Pw values", ylab="", main="L W W L W W W")

```

## 2M2  
Now assume a prior for Pw that is equal to zero when Pw < 0.5 and is a positive constant when Pw ≥ 0.5. Again compute and plot the grid approximate posterior distribution for each of the sets of observations in the problem just above.  

```{r}
prior <- ifelse(pw_grid < 0.5, 0, 1)

# posterior
posterior_1 <- get_posterior(prior, likelihood_1)
posterior_2 <- get_posterior(prior, likelihood_2)
posterior_3 <- get_posterior(prior, likelihood_3)


par(mfrow = c(1,3))
plot(posterior_1 ~ pw_grid, type="l", xlab="Pw values", ylab="probability", main="W W W")
plot(posterior_2 ~ pw_grid, type="l", xlab="Pw values", ylab="", main="W W W L")
plot(posterior_3 ~ pw_grid, type="l", xlab="Pw values", ylab="", main="L W W L W W W")

```

## 2M3  
Suppose there are two globes, one for Earth and one for Mars. The Earth globe is 70% covered in water. The Mars globe is 100% land. Further suppose that one of these globes—you don’t know which—was tossed in the air and produced a “land” observation. Assume that each globe was equally likely to be tossed. Show that the posterior probability that the globe was the Earth, conditional on seeing “land” (Pr(Earth|land)), is 0.23.  

```{r}
P_earth = 0.5
P_mars = 0.5
Pw_given_earth = 0.7
Pw_given_mars = 0.0

P_earth_given_land = ((1-Pw_given_earth)*P_earth) / ((1-Pw_given_earth)*P_earth + (1-Pw_given_mars)*P_mars)

print(round(P_earth_given_land, 3))
```


$$
P(E \mid \sim W) = \\
\frac{P(\sim W \mid E)P(E)}{P(\sim W)} = \\
\frac{P(\sim W \mid E)P(E)}{P(\sim W \mid E)P(E)+P(\sim W \mid M)P(M)} = \\
\frac{0.3 \times 0.5}{0.3 \times 0.5 + 1 \times 0.5} = \\
\frac{0.15}{0.65} \approx .231
$$

## 3E1 to 3E5

```{r}
n <- 1e3
p_grid <- seq(from=0, to=1, length.out=n)
prior <- rep(1, n)
likelihood <- dbinom(6, size=9, prob=p_grid)
posterior <- likelihood * prior / sum(likelihood * prior)

set.seed(100)
samples <- sample(p_grid, prob=posterior, size=1e4, replace = TRUE)

E1 <- sum(samples < 0.2) /1e4 
E2 <- sum(samples > 0.8) /1e4 
E3 <- sum(samples > 0.2 & samples < 0.8) /1e4 
E4 <- quantile(samples, 0.2)
E5 <- 1-quantile(samples, 0.8)
```

  
3E1. How much posterior probability lies below p = 0.2? **`r format(E1, scientific=F, digits=6)`**  

3E2. How much posterior probability lies above p = 0.8? **`r round(E2,3)`**   

3E3. How much posterior probability lies between p = 0.2 and p = 0.8? **`r round(E3,3)`**  

3E4. 20% of the posterior probability lies below which value of p? **`r round(E4,3)`**  

3E5. 20% of the posterior probability lies above which value of p? **`r round(E5,3)`**    

## 3E6    
Which values of p contain the narrowest interval equal to 66% of the posterior probability?  
```{r}
E6 <- rethinking::HPDI(samples, prob = 0.66)
E6
```

## 3E7  
Which values of p contain 66% of the posterior probability, assuming equal posterior probability both below and above the interval?  
```{r}
E7 <- rethinking::PI(samples, prob = 0.66)
E7
```

## 3M1  
Suppose the globe tossing data had turned out to be 8 water in 15 tosses. Construct the posterior distribution, using grid approximation. Use the same flat prior as before.  

```{r}
n <- 1e3
p_grid <- seq(from=0, to=1, length.out=n)
prior <- rep(1, n)
likelihood <- dbinom(8, size=15, prob=p_grid)
posterior <- likelihood * prior / sum(likelihood * prior)
```

## 3M2  
Draw 10,000 samples from the grid approximation from above. Then use the samples to calculate the 90% HPDI for p.  

```{r}
set.seed(100)
samples <- sample(p_grid, prob=posterior, size=1e4, replace = TRUE)
M2 <- rethinking::HPDI(samples, prob = 0.90)
M2
```

## 3M3  
Construct a posterior predictive check for this model and data. This means simulate the distribution of samples, averaging over the posterior uncertainty in p. What is the probability of observing 8 water in 15 tosses?  
```{r}
dummy_w <- rbinom(1e4, size=15, prob=samples)
M3 <- sum(dummy_w == 8) / length(dummy_w)
M3

```

## 3M4  
Using the posterior distribution constructed from the new (8/15) data, now calculate the prob- ability of observing 6 water in 9 tosses.  
```{r}
dummy_w <- rbinom(1e4, size=9, prob=samples)
M4 <- sum(dummy_w == 6) / length(dummy_w)
M4 

```

## 3M5  
Start over at 3M1, but now use a prior that is zero below p=0.5 and a constant above p=0.5.  This corresponds to prior information that a majority of the Earth’s surface is water. Repeat each problem above and compare the inferences. What difference does the better prior make? If it helps, compare inferences (using both priors) to the true value p = 0.7.  

```{r}
n <- 1e3
p_grid <- seq(from=0, to=1, length.out=n)
prior <- ifelse(p_grid<0.5, 0, 1)
likelihood <- dbinom(8, size=15, prob=p_grid)
posterior <- likelihood * prior / sum(likelihood * prior)

set.seed(100)
samples <- sample(p_grid, prob=posterior, size=1e4, replace = TRUE)
M2b <- rethinking::HPDI(samples, prob = 0.90)
M2b
M2
```

The narrowest (densest) interval using a prior <code> ifelse(p_grid<0.5, 0, 1) </code> is narrower than when using a prior of constant 1. In other words, starting with the second prior produces a narrower interval with the same amount of probability (i.e. 90%)--so under the second prior, $P_w$ is believed to be between 0.5 and 0.71, as opposed to between 0.33 and 0.72 with the same degree of probability (90%).  

```{r}
dummy_w <- rbinom(1e4, size=15, prob=samples)
M3b <- sum(dummy_w == 8) / length(dummy_w)
M3b
M3
```
Similarly, under the second prior (<code> ifelse(p_grid<0.5, 0, 1) </code>), the model assigns a slightly higher probability to observing 8 W in 15 tosses, than under the constant prior 1. (16.3% vs 15.7%)  

```{r}
dummy_w <- rbinom(1e4, size=9, prob=samples)
M4b <- sum(dummy_w == 6) / length(dummy_w)
M4b
M4
```

Lastly, under the second prior (<code> ifelse(p_grid<0.5, 0, 1) </code>), the model assigns a slightly higher probability to observing 6 W in 9 tosses, than under the constant prior 1. (23.4% vs 22.4%)

